---
title: "Bootylator: a bootstrap simulator for estimating survivals"
author: "Bobby Hsu"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Bootylator: a bootstrap simulator for estimating survivals}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, echo = FALSE, message = FALSE}
knitr::opts_chunk$set(collapse = T, comment = "#>")
library(bootylator)
```

## Background

The CSS/FPC bootstrap program (i.e. the BT series) was first developed by FPC biometrician Tom Berggren and programmer Henry Franzoni in the early 2000's. The BT programs were initially written in the FoxPro programming language, and later translated into C# by FPC programmer Sergei Rassk. Began in 2011, with the increasing popularity of R statistical software, FPC biometrician Jack Tuomikoski started developing a new version of bootstrap program to run simulations and process results using a suite of R functions (i.e. `bootstrapper`, `release`, and `DoCalcs`). As of 2018 CSS annual report, CSS/FPC had been conducting bootstrap simulations using both BT and Tuomikoski's bootstrap programs.

FPC continued to develop bootstrap program with two goals in mind. First, FPC would like to migrate all bootstrap programs to a single platform. Second, the bootstrap programs would preferably be written in R, so future updates and improvements could be done by FPC biologists or biometrician. In 2017, FPC developed `bootylator` package that bundled R functions to run CSS bootstraps and export results. `bootylator` retained all functionalities of BT and Tuomikoski's bootstrap programs, including weighted bootstrap. And `bootylator` added new capabilities such as finite population correction that would allow a more robust estimation with small sample size and avoid getting results with NA's. `bootylator` also included a function to simulate fake mark-recapture data for research purpose.

A basic workflow for `bootylator` to estimate CSS survivals:

* Loading and formatting data.

* Run bootstrap simulation.

* Output results.

Additional functionalities of `bootylator`:

* Calculate point estimates for survivals and detection without bootstrap.

* Generate fake mark-recapture data for simulation studies.


## Loading and Formatting Data

`format_dat()` formats the designated csv file before bootstrapping. The csv file has a standard format and is set by the FPC biologists. `format_dat()` flags and identifies fish that have questionable burnham histories (type "y" to see the list).

```{r}
dat_in<- format_dat(file_name='SR HCH 2016 CLWH_SU_new_format.csv', mig_yr=2016, wgt='n')
```


## Run the Bootstrap

`bootystrapper()` is the function that automates the bootstrap simulation. It takes the estimation function, `surv_cal()` or `mark_calc()`, and run it throung the iteration process.  User specifies number of iterations, options for incorporating weighted bootstrap, and finite population correction. Users can also specify to only get estimates for reach survival and detection (and omit other metrics). `surv_cal()` estimates using closed-form CJS, and `mark_calc()` calls package *RMARK* or *marked* to estimate survivals using logit link. Also, `bootystrapper()` shows a progress bar and prints processing time at the end.

```{r}
booty<- bootystrapper(dat_in, surv_calc, iter=10, wgt='n', wgt_init='n', fpc='y', phi_p_only='n')
```

To save bootstrap results in the SQL server (CSSOUTPUT), specify file name in "save_name." for example, `bootystrapper(dat_in, surv_calc, iter=10, save_name='SR HCH 2016 CLWH_SU')` indicates saving results under a string that includes "SR HCH 2016 CLWH_SU." Nothing would be save if "save_name" is not specified.


## Output

To summarize and export the results, use `doCalcs()` for Snake River or `doCalcs_uc()` for upper Columbia. Parameter outputs are appended in CSSREPORT in the SQL server and in csv file in current working directory. Exporting results is suppressed for the example below.

```{r}
ans<- doCalcs(booty, species='CH', reaches=6, makefile='n')
print(format(ans[,c(2,5,3)], scientific = F))
```


## Additional Functionalities

*Point Estimates for Survivals and Detection*

`surv_calc()` (or `mark_calc()`) can be used as a stand-alone function to calculate reach survivals and detection. Capture histories are the only information needed for an input file. Each capture event, including release, should have its own column. Specified number of occasion should match the number of column of capture history input file. Options for weighting (*wt* and *wt_i*) are not applicable, and should be specified as *"n"*. *phi_p_only* enables calculations only for survivals and detection.

```{r}
dat_in[1:5, 1:8]
pt_est<- surv_calc(dat_in, i=1, nocc=8, wt='n', wt_i='n', phi_p_only='y', fpc='y')
print(pt_est)
```

`marray()` can be used to extract m-array table.

```{r}
m_ij<- marray(dat_in, n_occ=8)
print(m_ij)
```

*Simulation Study*

`siml_cjs()` creates a fake detection history based on user specifications.

* *big_phi* sets reach survivals. it has *n_occ- 1* columns and *mrkd* rows.
* *big_p* sets detection probabilities for each reach. It has the same dimension as *big_phi*.
* *mrkd* indicates amount of fish released.
* *remv* indicates the portion of fish removed (transported or died).
* *n_occ* is the number of events/detection.
* *intgr* is the portion of "integrated" group. Integrated + Segregated = 1.
* *surv_diff* identifies the difference in survival between the integrated and segregated groups.
* *grp_t* is the portion of transported group. Group T + Group R = 1
* *adu_rtn* is the adult return rate.

```{r}
n_occ<- 8
mrkd<- 10
phi_real<- rep(0.9, n_occ-1) # set survival
p_real<- rep(0.45, n_occ-1) # set detection
remv<- 0.01 # portion removed at occ 2 and 3
intgr<- 0.23
surv_diff<- 0.07
grp_t<- 0.7
adu_rtn<- 0.05
big_phi<- matrix(phi_real, ncol= n_occ-1, nrow= mrkd)
big_p<- matrix(p_real, ncol= n_occ-1, nrow= mrkd)

ch<- siml_cjs(big_phi, big_p, mrkd, remv, n_occ, intgr, surv_diff, grp_t, adu_rtn)
ch
```





