---
title: "Bootylator: a bootstrap simulator for estimating survivals"
author: "Bobby Hsu"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Bootylator: a bootstrap simulator for estimating survivals}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, echo = FALSE, message = FALSE}
knitr::opts_chunk$set(collapse = T, comment = "#>")
library(bootylator)
```

## Background

The CSS/FPC bootstrap program (i.e. the BT series) was first developed by FPC biometrician Tom Berggren and programmer Henry Franzoni in the early 2000's. The BT programs were initially written in the FoxPro programming language, and later translated into C# by FPC programmer Sergei Rassk. Began in 2011, with the increasing popularity of R statistical software, FPC biometrician Jack Tuomikoski started developing a new version of bootstrap program to run simulations and process results using a suite of R functions (i.e. `bootstrapper`, `release`, and `DoCalcs`). As of 2018 CSS annual report, CSS/FPC had been conducting bootstrap simulations using both BT and Tuomikoski's bootstrap programs.

FPC continued to develop bootstrap program with two goals in mind. First, FPC would like to migrate all bootstrap programs to a single flatform. Second, the bootstrap programs would preferably be written in R, so future updates and improvements could be done by FPC biologists or biometrician. In 2017, FPC developed `bootylator` package that bundled R functions to run CSS bootstraps and export results. `bootylator` retained all functionalities of BT and Tuomikoski's bootstrap programs, including weighted bootstrap. And `bootylator` added new capabilities such as finite population correction that would allow a more robust estimation with small sample size and avoid getting results with NA's. `bootylator` also included a function to simulate fake mark-recapture data for research purpose.

The basic procedures for using `bootylator` to estimate CSS survivals are the following:

* Loading and formatting data.

* Run bootstrap simulation.

* Output results.

Additional functionalities of `bootylator`:

* Calculate point estimates for survivals and detection without bootstrap.

* Generate fake mark-recapture data for simulation studies.


## Loading and Formatting Data

`format_dat()` formats the desinated csv file before bootstrapping. The csv file should be set up in the same format as for BT series and Tuomikoski's bootstrap program. `format_dat()` flags and identifies fish that have questionable burnham histories (type "y" to see the list).

```{r}
dat<- format_dat(file_name='SR WCH 2017 AGCW.csv', wgt='n')
```


## Run the Bootstrap

Using `bootystrapper()`

## Output

Using `doCalcs_v3()` and `doCalcs_uc()`

## *Additional Functionalities*

## Point Estimates for Survivals and Detection

Using `surv_calc()` as a stand-alone

## Simulation Study

Using `siml_cjs()`

Vignettes are long form documentation commonly included in packages. Because they are part of the distribution of the package, they need to be as compact as possible. The `html_vignette` output type provides a custom style sheet (and tweaks some options) to ensure that the resulting html is as small as possible. The `html_vignette` format:

- Never uses retina figures
- Has a smaller default figure size
- Uses a custom CSS stylesheet instead of the default Twitter Bootstrap style

## Vignette Info

Note the various macros within the `vignette` section of the metadata block above. These are required in order to instruct R how to build the vignette. Note that you should change the `title` field and the `\VignetteIndexEntry` to match the title of your vignette.

## Styles

The `html_vignette` template includes a basic CSS theme. To override this theme you can specify your own CSS in the document metadata as follows:

    output: 
      rmarkdown::html_vignette:
        css: mystyles.css

## Figures

The figure sizes have been customised so that you can easily put two images side-by-side. 

```{r, fig.show='hold'}
plot(1:10)
plot(10:1)
```

You can enable figure captions by `fig_caption: yes` in YAML:

    output:
      rmarkdown::html_vignette:
        fig_caption: yes

Then you can use the chunk option `fig.cap = "Your figure caption."` in **knitr**.

## More Examples

You can write math expressions, e.g. $Y = X\beta + \epsilon$, footnotes^[A footnote here.], and tables, e.g. using `knitr::kable()`.

```{r, echo=FALSE, results='asis'}
knitr::kable(head(mtcars, 10))
```

Also a quote using `>`:

> "He who gives up [code] safety for [code] speed deserves neither."
([via](https://twitter.com/hadleywickham/status/504368538874703872))
